Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-23 paper.tex
experiments

The analysis of the traversability of tunnels by flexible ligands can be formulated as a motion planning problem and solved by sampling-based planners.
The sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands.
Although only the low-energy conformations are generated and saved to $\L$, the number of all conformations can still be large for
ligands with many rotatable bonds.
The aim of this experiment was to verify the procedure to select a subset of ligand conformations from $\L$.

Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller (packed) conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which  $|\L|=1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $|\L|=7560$ conformations.

@@

@@

The next studied case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine, which is a known inhibitor of the protein 4L2L.
We therefore assume that the ligand can reach the active site.
%which was in the protein crystal structure (PDB ID 4L2L). 
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial pdb structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels towards the active site (defined by residues 134 (Alanine) and 311 (Phenylalanine) for probe 0.9~\AA\ were calculated
using CAVER 3.0.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The detected the tunnels slightly differ due to conformational changes.
Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA with the average lengths 22.3~\AA, 25.3~\AA and 28.4~\AA respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
%The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
%The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
%in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
%Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).

%The most time-consuming part of the planners is the collision detection. 
%Although both methods can utilize the same number of collision detection queries in the expansion step, 
%\RB\ is slower than \RA, because  \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
%This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.

%The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
%it allows the ligands to retract more towards $\qrand$.
%On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
%Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
%However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
%By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.

The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
Due to low bottleneck of the tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.

Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
the planner prefers to apply the higher scales if possible.
%The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
%that leads to a new collision-free configuration.

The information about the scales used at the trajectory points 
brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
%To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
%The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.

>>>>

The analysis of the traversability of tunnels by flexible ligands can be formulated as a motion planning problem and solved by sampling-based planners.
The sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel,vonasek2016application}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands and eventhough only
low-enery conformations are stored in $\L$, the set can still contain thousands of conformations.
For a fast motion planning, we utilize a subset of the original set $\L$.
The aim of this experiment was to verify the procedure to select the subset.

Two selection strategies were tested:
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller (packed) conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which  $|\L|=1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $|\L|=7560$ conformations.

@@

The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial pdb structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0 for the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The detected the tunnels slightly differ due to conformational changes.
Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\ and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\ and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.

As the tested ligand is a known inhibitor of the protein 4L2L, we assume that the ligand can reach the active site.
We tested several scaling-down factors from $\smin=0.4$ to $\smin=0.6$.

Example of a found trajectory with the depicted final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
TODO: report num of travewrsable frames

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
%The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
%The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
%in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
%Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
%The most time-consuming part of the planners is the collision detection. 
%Although both methods can utilize the same number of collision detection queries in the expansion step, 
%\RB\ is slower than \RA, because  \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
%This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
%The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
%it allows the ligands to retract more towards $\qrand$.
%On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
%Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
%However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
%By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.

The presented work is motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
Due to low bottleneck of the tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.

Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
the planner prefers to apply the higher scales if possible.

{\color{red}
    Neni to blbost? (ta prvni veta)
Determination of the proper scaling-down factor is beyond the scope of this paper, where we focus on the algorithmic part of the problem.
The computed trajectories are further visualized and analyzed.
The scales used at the trajectory points brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
Beyond this point, the ligand can move at larger scales again.
}

As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}. 
The necessity to scaled-down even more than 0.8, which is a value used by other tools~\cite{cortes2010simulating}, was
shown also in the second experiment, where the trajectory for inhibitor of the 4L2L protein required scaled-down to 0.5 or 0.6.

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
The ligand shape is taken into account and the ligand flexibility is modeled using a predefined set of its conformations.
The ligand conformations are prepared using the Rosetta tool, ranked by their energy and only low-energy conformations are used in the planning.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
%To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
%The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
